---
title: "Exploiting the Depmap cancer dependency data using the depmap R package"
author:
- name: Theo Killian
  affiliation: Computational Biology, UCLouvain
- name: Laurent Gatto
  affiliation: Computational Biology, UCLouvain
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{depmap use cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
# http://bioconductor.org/packages/release/bioc/vignettes/BiocWorkflowTools/inst/doc/Generate_F1000_Latex.html
```

# Abstract

The `depmap` package facilitates access in the R environment to the data from
the Depmap project, a multi-year collaborative effort by the Broad Institute and
Wellcome Sanger Institute, mapping genetic and chemical dependencies and other
molecular biological measurements of over 1700 cancer cell lines. The `depmap`
package formats this data for use of popular R data analysis and visualizing
tools such as `dplyr` and `ggplot2` to represent and visualize these rich
datasets. In addition, the depmap package utilizes `ExperimentHub`, storing
versions of the Depmap data accessible from the Cloud, which may be selectively
downloaded, providing a reproducible research framework to support exploiting
this data. This paper describes a workflow demonstrating how to access and
visualize the Depmap data in R using this package.

# Keywords

cancer, cancer dependency, Depmap, R, dplyr, ggplot2, ExperimentHub, Cloud, data
mining, reproducibility, Bioconductor

# Introduction

The consequences of genomic alterations of cancer cells on the molecular
biological landscape of the cell may result in differential vulnerabilities, or
"dependencies" compared to those of healthy cells. An example may be a gene not
necessary for the survival in healthy cells, but essential for the vitality in a
particular cancer cell line. The exact nature of many of these dependencies in
cancer cell lines is not completely understood. A map illustrating the
relationships between the genetic features of cancer and cancer dependencies is
desirable. The Cancer Dependency Map or "Depmap", a strategic collaboration
between the Broad Institute and the Wellcome Sanger Institute, aims to map such
dependencies in a broad range cancer cell lines, with the intention of
exploiting this knowledge to develop new therapies in precision cancer medicine.

As of the 2019 Q3 Depmap release, 1736 human cancer cell lines have been mapped
for dependencies. The sampling of tested cell lines is intended to mirror the
distribution of human cancer. Information derived from the testing of new cell
lines in the process of being mapped is added to subsequent quarterly dataset
releases. The primary methods utilized in mapping genomic dependencies are gene
knockout and gene knockdown, performed been CRISPR and RNAi, respectively.
Genetic dependency is calculated from the measured log fold change in the of
amount of shRNA detected after a fixed period after gene knockout or knockdown.
To correct for potential off-target effects of gene knockout or knockdown in
overestimating dependency with RNAi and CRISPR, the Depmap study utilized two
algorithms, DEMETER2 and CERES, to moderate the dependency estimation.
Measurements of chemical dependencies were also undertaken with the Depmap PRISM
viability screens that, as of the 2019 Q3 release, consisted of 4,518 compounds
tested against 578 cancer cell lines. The Depmap project has also compiled other
datasets of further molecular biological characterization of cancer cell lines,
such as genomic copy number, Reverse Phase Protein Array (RPPA) data, TPM gene
expression data for protein coding genes and metadata containing mutation calls.
These datasets are updated quarterly on a release schedule and are publically
available under [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) licence.

There is clearly a great wealth of information relevant to cancer research to be
found within the data generated by the Depmap project. Users who wish to begin
investigation of this data may start by using the [data explorer](https://depmap.org/portal/interactive/)
currently available on the Depmap website. However, this data viewer does not
allow the visualization of more than 2 genes at a time, nor affords viewing data
from older datasets. Therefore, in order to dig deeper into this data and
perform more complex analyses, such as clustering of genes or cell lines on
dependency, or global comparisons of multiple features, a more sophisticated,
customizable data analysis tool is required. Moreover, even though the Depmap
datasets have been made available for public access, it may be difficult for
those not skilled in data manipulation to efficiently import and visualize this
data, especially in regards to integrating data together from multiple datasets.
For example, the Depmap project created a unique private key corresponding to
each cancer cell line, but this value is only included in the metadata dataset
for each release. Another difficulty in fully exploiting this data is that some
datasets are not of the same format, with some datasets having cancer cell lines
as columns and others listing them as rows and vice versa.

To address the issues discussed above, the `depmap` Bioconducter package was
created. The Depmap datasets have been converted to possess the private key
"depmap_id" for all data tables, in order to make features more comparable. In
addition, the Depmap datasets were converted to long format, to facilitate the
use of common R packages such as `dplyr` and `ggplot2` in the R statistical
environment. By creating visualizations in the R environment, aides in
reproducible research, as compared to using the Depmap data viewer, which is a
black box. 

As new Depmap datasets are released on a quarterly basis, it is not feasible to
include all dataset files directly within the directory of the R package. To
keep the package lightweight, the `depmap` package `ExperimentHub` was used to
store and retrieve all versions (starting from 19Q1) of the Depmap data in the
Cloud using AWS. The `depmap` package contains accessor funtions to directly
download from the Cloud the most current datasets into the R environment.
Specific datasets, such as older datasets which have been used in prior research
can also be similarly downloaded. This feature has the added advantage of
enhancing reproducible research, such that specific versions of Depmap data can
be selected, in addition to having access to the most current datasets. The
`depmap` R package is available as part of Bioconductor at:
https://bioconductor.org/packages/devel/data/experiment/html/depmap.html

# Use cases

The features of primary interest from the Depmap Project are the measurements of
cancer gene dependency scores, found in datasets `rnai` and `crispr`. This score
is an expression of how vital a particular gene for a given cancer cell line is
in terms of the lethality resulting from the knockout/knockdown of that gene.
For example, a highly negative dependency score is derived from a large negative
log fold change in the population of cancer cells after gene knockout/knockdow
and implies that cell line is highly dependent on that gene.

```{r load_libraries, message=FALSE, warning=FALSE, echo=TRUE}
library("dplyr")
library("ggplot2")
library("viridis")
library("tibble")
library("gridExtra")
library("stringr")
library("depmap")
library("ExperimentHub")
library("plotly")
```

Load the `rnai`, `crispr` and `copyNumber` datasets for visualization from the
`19_Q1` release.

```{r load_data, message=FALSE, warning=FALSE, echo=TRUE}
## create ExperimentHub query object
eh <- ExperimentHub()
query(eh, "depmap")

  # EH3080 | rnai_19Q3            
  # EH3081 | crispr_19Q3          
  # EH3082 | copyNumber_19Q3      
  # EH3083 | RPPA_19Q3            
  # EH3084 | TPM_19Q3             
  # EH3085 | mutationCalls_19Q3   
  # EH3086 | metadata_19Q3        
  # EH3087 | drug_sensitivity_19Q3

## note: the datasets listed above are from the 19Q1 release. Newer datasets,
## such as 19Q2 and 19Q3 are available.
rnai <- eh[["EH3080"]]
#crispr <- eh[["EH3081"]]
copyNumber <- eh[["EH3082"]]
TPM <- eh[["EH3084"]]
# RPPA <- eh[["EH3083"]]
mutationCalls <- eh[["EH3085"]]
metadata <- eh[["EH3086"]]
# EH3087 | drug_sensitivity_19Q3
```

```{r message=FALSE, warning=FALSE}
# Basic histogram
rnai %>% select(gene, gene_name, dependency) %>% 
         filter(gene_name == "PIK3CA") %>% 
         ggplot(aes(x=dependency)) + geom_histogram() +
         geom_vline(xintercept=mean(rnai$dependency, na.rm = TRUE),
                    linetype = "dotted", color = "red") +
         ggtitle("Histogram of dependency scores for gene PIK3CA")
```

```{r message=FALSE, warning=FALSE}
meta_rnai <- metadata %>%
  select(depmap_id, lineage) %>%
  full_join(rnai, by = "depmap_id") %>%
  filter(gene_name == "PIK3CA") %>% 
  full_join((mutationCalls %>% select(depmap_id, entrez_id, is_cosmic_hotspot, var_annotation)),
                      by = c("depmap_id", "entrez_id")) # %>% 
  #full_join((TPM %>% select(depmap_id, gene, expression)),
  #                    by = c("depmap_id", "gene"))   

p1 <- meta_rnai %>% ggplot(aes(x=dependency, y=lineage, #text = paste("expression:", expression)
                               )) +
  geom_point(alpha = 0.4, size = 0.5) +
  geom_point(data = subset(meta_rnai, var_annotation == "damaging"), color = "red") +
  geom_point(data = subset(meta_rnai, var_annotation == "other non-conserving"), color = "blue") +
  geom_point(data = subset(meta_rnai, var_annotation == "other conserving"), color = "cyan") +
  geom_point(data = subset(meta_rnai, is_cosmic_hotspot == TRUE), color = "orange") +
  geom_vline(xintercept=mean(meta_rnai$dependency, na.rm = TRUE), linetype = "dotted", color = "red") +
  ggtitle("Scatterplot of dependency scores for gene PIK3CA by lineage")

ggplotly(p1)
```

```{r}
rm(p1)
rm(meta_rnai)
```

```{r}
meta_tpm <- metadata %>%
            select(depmap_id, lineage) %>%
            full_join(TPM, by = "depmap_id") %>%
            filter(gene_name == "PIK3CA")

p2 <- meta_tpm %>%
      ggplot(aes(x=lineage, y=expression, fill = lineage)) +
      geom_boxplot(outlier.alpha = 0.1) +
      ggtitle("Boxplot of expression values for gene PIK3CA by lineage")
p2 <- p2 + theme(axis.text.x = element_text(angle = 45))
ggplotly(p2)
```

```{r}
rm(p2)
rm(meta_tpm)
```

Let's try this again with RNAI

```{r}
# expression vs rnai gene dependency for Rhabdomyosarcoma Sarcoma
sarcoma <- metadata %>%
           select(depmap_id, cell_line, primary_disease, subtype_disease) %>%
           filter(primary_disease == "Sarcoma",
                  subtype_disease == "Rhabdomyosarcoma")
rnai_sub <- rnai %>% dplyr::select(depmap_id, gene, gene_name, dependency)
tpm_sub <- TPM %>% dplyr::select(depmap_id, gene, gene_name, expression)
sarcoma_dep <- sarcoma %>%
               left_join(rnai_sub, by = "depmap_id") %>%
               select(-cell_line, -primary_disease, -subtype_disease, -gene_name)
sarcoma_exp <- sarcoma %>% left_join(tpm_sub, by = "depmap_id")
sarcoma_dat_exp <- full_join(sarcoma_dep, sarcoma_exp,
                             by = c("depmap_id", "gene"))

p3 <- ggplot(data=sarcoma_dat_exp, aes(x=dependency, y=expression,
                                       text = paste("gene:", gene_name))) +
      geom_point(alpha = 0.4, size = 0.5) +
      geom_vline(xintercept=mean(sarcoma_dat_exp$dependency, na.rm = TRUE), linetype = "dotted", color = "red") +
      geom_hline(yintercept=mean(sarcoma_dat_exp$expression, na.rm = TRUE), linetype = "dotted", color = "red") +
      ggtitle("Scatterplot of rnai dependency vs expression values for gene PIK3CA")  
ggplotly(p3)
```

```{r}
rm(p3)
rm(sarcoma)
rm(rnai_sub)
rm(tpm_sub)
rm(sarcoma_dep)
rm(sarcoma_exp)
rm(sarcoma_dat_exp)
```

```{r}
# copy number by lineage
meta_copy <- metadata %>%
            select(depmap_id, lineage) %>%
            full_join(copyNumber, by = "depmap_id") %>%
            filter(gene_name == "PIK3CA")

p4 <- meta_copy %>%
      ggplot(aes(x=lineage, y=log_copy_number, fill = lineage)) +
      geom_boxplot(outlier.alpha = 0.1) +
      ggtitle("Boxplot of log copy number for gene PIK3CA by lineage")
p4 <- p4 + theme(axis.text.x = element_text(angle = 45))
ggplotly(p4)
```


## Find dependency score for "BRCA1" on "184A1_Breast"

We will demonstrate how to obtain individual dependency scores corresponding to
a specific gene and cell lineage. For example, shown below is the dependency of
a breast cancer lineage, such as `184A1_BREAST` has on a human tumor suppressor
gene, like `BRCA1` when it is knocked down via rnai. Shown below is the
comparison for data found within the `rnai` dataset. This shows a score which is
slightly positive, indicating that the knockdown of this gene is slightly
beneficial to the vitality of this cancer cell lineage. However, it may be
insightful to put this single dependency score in context.

```{r dep_score_BRCA1_184A1Breast, echo=TRUE}
# dep_score_BRCA1_184A1Breast <- rnai %>%
#                                select(cell_line, gene_name, dependency) %>%
#                                filter(cell_line == "184A1_BREAST",
#                                       gene_name == "BRCA1")
# 
# dep_score_BRCA1_184A1Breast
```

## Average gene dependency for "BRCA1"

Shown below is the average dependency score for `BRCA1` for all cancer cell
lines in the `rnai` dataset.

```{r, BRCA1_Avg_Dep_Score, echo=TRUE}
# brca1_dep_score_avg_rnai <- rnai %>%
#                             select(gene_name, dependency) %>%
#                             filter(gene_name == "BRCA1") %>%
#                             summarise(mean_dependency_brca1 =
#                                       mean(dependency, na.rm=TRUE))
# 
# brca1_dep_score_avg_rnai
```

## Average gene dependency for all genes in the `rnai` dataset


Or to see the average gene dependency across all genes in the entire `rnai`
dataset. As one can see below, the average dependency for an average gene in the
`rnai` dataset is slightly negative but close to zero.

```{r, all_gene_ds_avg_rnai, echo=TRUE}
# all_gene_dep_score_avg_rnai <- rnai %>%
#                                select(gene_name, dependency) %>%
#                                summarise(mean_dependency_all_genes_rnai =
#                                          mean(dependency, na.rm=TRUE))
# all_gene_dep_score_avg_rnai
```

## Cell lines in the `rnai` dataset with "soft tissue" in the name

If we are interested researching soft tissue sarcomas and wanted to find the
cell lines withing the `rnai` dataset that had "soft tissue" in the CCLE name of
cancer cell line, and sort by the highest dependency score. The results of such
a search is shown below. Note: CCLE names are in ALL CAPS with an underscore.

```{r, soft_tissue_cell_lines, echo=TRUE}
# soft_tissue_dep_rnai <- rnai %>%
#                         select(cell_line, gene_name, dependency) %>%
#                         filter(stringr::str_detect(cell_line, "SOFT_TISSUE")) %>%
#                         arrange(dependency)
# 
# soft_tissue_dep_rnai
```

## Cell lines with dependency for a  entrez_id of interest

Sometimes it is difficult to find the subset with the exact gene name one wishes
to find. In this case, it is better to search by `entrez_id`. For example, a
[recent paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6251792/) describes
gene knockdown of *NRF2* increases chemosensitivity in certain types of cancer.
It might be interesting to see what interactions knockdown of this gene has on
other cancer cell lines. However, searching by *filter(gene_name == "NRF2")*
will not yield any results. We know from NCBI that the Entrez ID for this gene
is "4780" and it is possible to search this dataset by that criteria. Here it
can be shown that the gene name for *NRF2* in the `rnai` dataset is *NFE2L2*.

```{r cell_lines_with_entrez_id_NRF2, echo=TRUE}
# entrez_id_NRF2 <- rnai %>%
#                   select(entrez_id, cell_line, gene_name, dependency) %>%
#                   filter(entrez_id == "4780")
# 
# entrez_id_NRF2
```

## Cell lines with dependency for "NFE2L2"

Below the highest dependency scores via rnai knock down of a specific gene,
*NFE2L2* will be obtained and the cancer cell lines associated with those values
will be listed. It appears that the knockdown of this gene is strongly
associated with cell death with in lung and kidney cancer cell lines.

```{r greatest_Dep_Score_NFE2L2, echo=TRUE}
# top_dep_score_NFE2L2_rnai <- rnai %>%
#                              select(cell_line, gene_name, dependency) %>%
#                              filter(gene_name == "NFE2L2") %>%
#                              arrange(dependency)
# 
# top_dep_score_NFE2L2_rnai
```

## Genes for cell line "NCIH2066_LUNG"

If we would like to obtain the top 10 lowest dependency scores for a
particular cell line (for example `NCIH2066_LUNG`) along with the genes
associated with those values:

```{r top_dep_score_NCIH2066_LUNG_rnai, echo=TRUE}
# top_dep_score_NCIH2066_LUNG_rnai <- rnai %>%
#                                 select(cell_line, gene_name, dependency) %>%
#                                 filter(cell_line == "NCIH2066_LUNG") %>%
#                                 arrange(dependency)
# 
# top_dep_score_NCIH2066_LUNG_rnai
```

## Most and least RNAi dependency genes

Below shows the most significant genes that deplete cancer cell lines upon
knockdown and their dependency scores for the entire `rnai` data.

```{r greatest_dep_score_gene_rnai, echo=TRUE}
# greatest_dep_score_gene_rnai <- rnai %>%
#                             select(cell_line, gene_name, dependency) %>%
#                             arrange(dependency)
# 
# greatest_dep_score_gene_rnai
```

Below shows the least significant genes that induce cancer cell line vitality
upon knockdown and their dependency scores for the entire `rnai` data.
Unsurprisingly, we see high incidence of "TP53", a well known cancer driver.

```{r cell_line_gene_rnai_lowest_dep_score, echo=TRUE}
# lowest_dep_score_gene_rnai <- rnai %>%
#                             select(cell_line, gene_name, dependency) %>%
#                             arrange(desc(dependency))
# 
# lowest_dep_score_gene_rnai
```

## Most and least CRISPR-Cas9 dependency genes

Below we will apply some of the same selections as shown in the above examples
on the `crispr` gene knockout dataset and observe the difference between that
dataset and `rnai`. First we will look at the most significant dependency scores
in the `crispr` dataset. As can be seen below, there is a different population
of significant genes with the highest dependency score.

```{r cell_line_gene_crispr_greatest_dep_score, echo=TRUE}
# greatest_dep_score_gene_crispr <- crispr %>%
#                                 select(cell_line, gene_name, dependency) %>%
#                                 arrange(dependency)
# 
# greatest_dep_score_gene_crispr
```

First we will look at the least significant (most cancer inducing) dependency
scores in the `crispr` dataset.

```{r cell_line_gene_crispr_lowest_dep_score, echo=TRUE}
# lowest_dep_score_gene_crispr <- crispr %>%
#                             select(cell_line, gene_name, dependency) %>%
#                             arrange(desc(dependency))
# 
# lowest_dep_score_gene_crispr
```

## Differences in RNAi and CRISPR-Cas9 dependency scores

Here we will plot the difference in expression between the most signficant genes
found in the `crispr` and `rnai` datasets.

```{r comparison_rnai_crispr_dep_scores, fig.height=6, fig.width=7, fig.align="center", echo=FALSE, message=FALSE}
# sort `crispr` dep scores by most cancer inducing
# top_20_dep_scores_crispr <- crispr %>%
#                                 select(gene_name, dependency) %>%
#                                 arrange(desc(dependency)) %>%
#                                 top_n(20)
# 
# # sort `crispr` dep scores by least cancer inducing
# lowest_20_dep_scores_crispr <- crispr %>%
#                                 select(gene_name, dependency) %>%
#                                 arrange(dependency) %>%
#                                 top_n(-20)
# 
# # sort `rnai` dep scores by most cancer inducing
# top_20_dep_scores_rnai <- rnai %>%
#                             select(gene_name, dependency) %>%
#                             arrange(desc(dependency)) %>%
#                             top_n(20)
# 
# # shorten gene name so it fits on plot
# top_20_dep_scores_rnai[12, 1] <- "GAS6-AS2"
# 
# # sort `rnai` dep scores by least cancer inducing
# lowest_20_dep_scores_rnai <- rnai %>%
#                                 select(gene_name, dependency) %>%
#                                 arrange(dependency) %>%
#                                 top_n(-20)
# 
# # rnai highest dep scores
# p1 <- ggplot(lowest_20_dep_scores_rnai) +
#         aes(gene_name, dependency, color = dependency) +
#         geom_point() + scale_colour_viridis() +
#         theme(axis.text.x = element_text(angle=90, hjust=1)) +
#         ggtitle("High DS genes rnai")
# 
# # crispr highest dep scores
# p2 <- ggplot(lowest_20_dep_scores_crispr) +
#         aes(gene_name, dependency, color = dependency) +
#         geom_point() + scale_colour_viridis() +
#         theme(axis.text.x = element_text(angle=90, hjust=1)) +
#         ggtitle("High DS genes crispr ")
# 
# # rnai lowest dep scores
# p3 <- ggplot(top_20_dep_scores_rnai) +
#         aes(gene_name, dependency, color = dependency) +
#         geom_point() + scale_colour_viridis() +
#         theme(axis.text.x = element_text(angle=90, hjust=1)) +
#         ggtitle("Low DS genes rnai")
# 
# # crispr lowest dep scores
# p4 <- ggplot(top_20_dep_scores_crispr) +
#         aes(gene_name, dependency, color = dependency) +
#         geom_point() + scale_colour_viridis() +
#         theme(axis.text.x = element_text(angle=90, hjust=1)) +
#         ggtitle("Low DS genes crispr")
# 
# #plot as 1x2 grid
# grid.arrange(p1, p2, p3, p4, nrow=2,
#              top = "Most Extreme Dep Scores for CRISPR and RNAI")
```

Compare the count of top 50 unique genes for `crispr` and `rnai` datasets for
the most cancer-vitality inducing genes.

```{r count_comp_rnai_crispr_dep_scores, fig.height=6, fig.width=7, fig.align="center", echo=FALSE, message=FALSE}
# get counts of top 50 genes in `crispr` that are most cancer-vitality inducing
# unique_lowest_dep_scores_gene_crispr <- crispr %>%
#                                     select(gene_name, dependency) %>%
#                                     arrange(desc(dependency)) %>%
#                                     top_n(50) %>%
#                                     count(gene_name) %>%
#                                     arrange(desc(n))
# 
# # TBC1D3 appears to be an extremely common for `crispr` and dominates top 100
# # most cancer-vitality-inducing genes.
# 
# # get counts of top 50 genes in `rnai` that are most cancer-vitality inducing
# unique_lowest_dep_scores_gene_rnai <- rnai %>%
#                                 select(gene_name, dependency) %>%
#                                 arrange(desc(dependency)) %>%
#                                 top_n(50) %>%
#                                 count(gene_name) %>%
#                                 arrange(desc(n))
# 
# # Whereas for `rnai` UBBP4 and ACTG1P4 are most common.
# 
# # shorten gene name to fit on plot
# unique_lowest_dep_scores_gene_rnai[9, 1] <- "GAS6-AS2"
# 
# # get counts of top 50 genes in `crispr` with greatest dependency
# unique_top_dep_scores_crispr <- crispr %>%
#                                 select(gene_name, dependency) %>%
#                                 arrange(dependency) %>%
#                                 top_n(-50) %>%
#                                 count(gene_name) %>%
#                                 arrange(desc(n))
# 
# # Most common gene for top 100 most dependent genes in `crispr` is RAN
# 
# # get counts of top 50 genes in `rnai` with greatest dependency
# unique_top_dep_scores_rnai <- rnai %>%
#                                 select(gene_name, dependency) %>%
#                                 arrange(dependency) %>%
#                                 top_n(-50) %>%
#                                 count(gene_name) %>%
#                                 arrange(desc(n))
# 
# # Most common genes for top 100 most dependent genes in `rnai` are RPL7, EIF3B
# # and RPL14.
# 
# # rnai highest dep scores
# p5 <- ggplot(unique_top_dep_scores_crispr, aes(x=gene_name, y=n)) +
#     geom_bar(stat='identity', fill="steelblue2") +
#     ylab("count") + theme(axis.text.x = element_text(angle=90, hjust=1)) +
#     ggtitle("Highest DS genes crispr")
# 
# p6 <- ggplot(unique_top_dep_scores_rnai, aes(x=gene_name, y=n)) +
#     geom_bar(stat='identity', fill="steelblue2") +
#     ylab("count") + theme(axis.text.x = element_text(angle=90, hjust=1)) +
#     ggtitle("Highest DS genes rnai")
# 
# p7 <- ggplot(unique_lowest_dep_scores_gene_crispr, aes(x=gene_name, y=n)) +
#     geom_bar(stat='identity', fill="steelblue2") +
#     ylab("count") + theme(axis.text.x = element_text(angle=90, hjust=1)) +
#     ggtitle("Lowest DS genes crispr")
# 
# p8 <- ggplot(unique_lowest_dep_scores_gene_rnai, aes(x=gene_name, y=n)) +
#     geom_bar(stat='identity', fill="steelblue2") +
#     ylab("count") + theme(axis.text.x = element_text(angle=90, hjust=1)) +
#     ggtitle("Lowest DS genes rnai")
# 
# ## plot as 1x2 grid
# grid.arrange(p5, p6, p7, p8, nrow=2,
#              top = "Top 50 CRISPR and RNAI genes with High and Low Dep Score")
```

Mean log copy number (total dataset) and mean log copy number for each gene

```{r mean_log_copy_num_gene, fig.height=4, fig.width=6, fig.align="center", echo=FALSE, message=FALSE}
# mean log copy number for all genes
# mean_log_copy_num_gene <- copyNumber %>%
#                             select(gene_name, log_copy_number) %>%
#                             summarise(mean_log_copy_number_all_genes =
#                                           mean(log_copy_number, na.rm = TRUE))
# 
# # get average value
# val_mean_log_copy_num_gene <- as.numeric(mean_log_copy_num_gene[1,1])
# 
# # mean log copy number for each gene
# each_log_copy_num_gene <- copyNumber %>%
#                             select(gene_name, log_copy_number) %>%
#                             group_by(gene_name) %>%
#                             summarise(mean_log_copy_number =
#                                           mean(log_copy_number, na.rm = TRUE))
# 
# # add an ID column
# all_log_copy_num_gene <-
#     as.data.frame(na.omit(each_log_copy_num_gene$mean_log_copy_number)) %>%
#     tibble::rowid_to_column(., "ID")
# 
# # add col names
# colnames(all_log_copy_num_gene, do.NULL = FALSE)
# colnames(all_log_copy_num_gene) <- c("gene", "log_copy_number")
# 
# # plot of mean copy number for every gene
# p9 <- ggplot(all_log_copy_num_gene) +
#             aes(x=gene, y=log_copy_number, color = log_copy_number) +
#             geom_point() +  scale_colour_viridis() +
#             geom_hline(yintercept = val_mean_log_copy_num_gene,
#             linetype = "dashed", color = "red") +
#             theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#             ggtitle("Mean log copy number change for every gene")
# p9
```

## Genes with greatest mean log copy number

Find genes with greatest mean log copy number

```{r greatest_log_copy_num_genes, fig.height=4, fig.width=6, fig.align="center", echo=FALSE, message=FALSE}
# mean log copy number for each gene
# greatest_log_copy_num_genes <- copyNumber %>%
#                                 select(gene_name, log_copy_number) %>%
#                                 group_by(gene_name) %>%
#                                 summarise(mean_long_copy_num =
#                                               mean(log_copy_number))%>%
#                                 arrange(desc(mean_long_copy_num)) %>%
#                                 top_n(20)
# 
# p10 <- ggplot(greatest_log_copy_num_genes, aes(x=gene_name,
#                                               y=mean_long_copy_num)) +
#             geom_bar(stat='identity', fill="steelblue2") +
#             ylab("log Copy Number") +
#             theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#             ggtitle("Top 20 genes with largest log copy number")
# p10
```

# Discussion and Outlook

We hope that this package will be used by cancer researchers to dig deeper into
the Depmap data and to support their research.

The depmap R package will continue to be maintained in line with the biannual
Bioconductor release, in addition to quarterly releases of Depmap data.

We welcome feedback and questions from the community. We also highly appreciate
contributions to the code in the form of pull requests.

# Software availability

Please note that this manuscript uses version 1.10 of the depmap R package,
which is in the development version of Bioconductor (v.3.10). There are two
possible ways of installing it:

1) Install the development version of R (v.3.6) — required for Bioconductor
v.3.10 — and install depmap using the command:

```{r}
# BiocManager::install('depmap', version = 'devel')
```

–or–
    
2) Install depmap 1.0 directly from the github repository using the devtools R
package:

```{r}
# install.packages('devtools')
# library(devtools)
# install_github('uclouvain-cbio/depmap')
```

# Session information

```{r echo = FALSE}
sessionInfo()
```

# Acknowledgements

# References

1. [Tsherniak, Aviad, et al. "Defining a cancer dependency map." Cell 170.3 (2017):
564-576.](https://www.ncbi.nlm.nih.gov/pubmed/28753430). 

2. Depmap, A Cancer Dependency Map to systematically identify genetic and
pharmacologic dependencies and the biomarkers that predict them. 2019. Retrieved
from https://depmap.org/portal/download/

3. McFarland, J. M., Ho, Z. V., Kugener, G., Dempster, J. M., Montgomery, P. G.,
Bryan, J. G., ... & Golub, T. R. (2018). Improved estimation of cancer
dependencies from large-scale RNAi screens using model-based normalization and
data integration. Nature communications, 9.

4. Meyers, R. M., Bryan, J. G., McFarland, J. M., Weir, B. A., Sizemore, A. E.,
Xu, H., ... & Goodale, A. (2017). Computational correction of copy number effect
improves specificity of CRISPR–Cas9 essentiality screens in cancer cells. Nature
genetics, 49(12), 1779.

5. depmap. Cancer Dependency Map Data Package. 2019. Retrieved
from https://bioconductor.org/packages/devel/data/experiment/html/depmap.html